---
title: role of life history traits and introduction effort in understanding invasion success
author: "Annemarie van der Marel"
date: "12/08/2020"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

Exploring the role of life history traits and introduction effort in understanding invasion success in mammals: A case study of Barbary ground squirrels
Annemarie van der Marel a, Jane M. Waterman a, Marta López-Darias b 
a Department of Biological Sciences, University of Manitoba, Winnipeg, Canada
b Island Ecology and Evolution Research Group, Instituto de Productos Naturales y Agrobiología (IPNA-CSIC), La Laguna, Tenerife, Canary Islands, Spain


AIM 1: we investigated life history traits of the Barbary ground squirrel, _Atlantoxerus getulus_, a species with a remarkablye low introduction effort (Machado, 1979), which will contribute to our understanding of invasion success when introduction efforts are minimal and will be essential to plan for future conservation programs on this species. 

AIM 2: we studied whether the Barbary ground squirrel's exceptional invasion success (despite low introduction effort) is due to a very fast life history profile through a comparison of these traits to other successfully invaded mammals using the dataset provided by Capellini et al. (2015). 

AIM 3: we examined whether number of founders and/or a fast life history influences invasion success of squirrels using the dataset provided by Bertolino (2009).  

# Load libraries
```{r, include=FALSE}
library(tidyverse)
library(ggplot2)
library(effsize)
library(survival)
library(survminer)
library(dplyr)
library(janitor)
library(ggpubr)
library(lme4)
library(MuMIn)
library(piecewiseSEM)
library(adephylo)
library(beanplot)
library(ggpubr)
```


# AIM 1: life history traits Barbary ground squirrels

For the following measurements, the data are available upon reasonable request from the authors:
 - Duration breeding season
 - duration that males are scrotal
 - Gestation length
 - Lactation length
 - Weaning length
 - adult body mass

## age at first birth
```{r}
age<-read.csv("age at first birth.csv", sep = ";")
head(age)

# model diagnostics
ggplot(data = age, aes(sample = age)) +
  stat_qq() +
  stat_qq_line()

ggplot(age, aes(x = fromdate, y = age)) +
  geom_point() +
  stat_smooth(method = "lm")
m <- lm(age~fromdate, data = age)
summary(m)
d <- data.frame(residuals = residuals(m),         # Residuals
                std_residuals = rstudent(m), # Studentized Residuals
                fitted = fitted(m),          # Fitted values
                cooks = cooks.distance(m))   # Cook's D
d <- mutate(d, observation = 1:nrow(d))           # Observation number

# histogram of residuals
ggplot(data = d, aes(x = std_residuals)) +
  geom_histogram(bins = 10)
# Check for Normality: QQ Normality plot of residuals
ggplot(data = d, aes(sample = std_residuals)) +
  stat_qq() +
  stat_qq_line()
# Check heteroscedasticity or constant variance
ggplot(d, aes(x = fitted, y = std_residuals)) +
  geom_point() +
  geom_hline(yintercept = 0)
# Cook's D
ggplot(d, aes(x = observation, y = cooks)) +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_hline(yintercept = 4/nrow(d), linetype = "dashed")
# outliers
d_outliers <- filter(d, cooks > 4/nrow(d))
d_outliers
age[c(1), ] # keep
age[c(8), ] # keep
age[c(18), ] # keep
# opt for nonparametric test

wilcox.test(age~ fromdate, data=age) # all females

# females with known date of birth/emergence
age1<-dplyr::filter(age, certainty=="certain")
wilcox.test(age~ fromdate, data=age1)

# mean and SE
agesum<-dplyr::select(age1, fromdate, age) %>%
  group_by(fromdate) %>%
  summarise_all(list(mean, sd, n)) %>%
  mutate(se=sd/sqrt(n))

mean(age1$age)
sd(age1$age)/sqrt(length(age1$age))

agesum
# clean environment
rm(age1, d_outliers, d, age, m )
```

## Maternal success rate 
Proportion of females that successfully raised their litter to emergence

```{r}
success<-c(28, 44, 63)
nonsuccess<-c(11, (58-44), (80-63))
rows<-3
m<-matrix(c(success, nonsuccess),nrow=rows)
rownames(m) = c("2014", "2015", "2016")     #  Naming the rows and
colnames(m) = c("success", "nonsuccess")    #  columns is optional
m

chisq.test(m)

# clean environment
rm(success, nonsuccess, rows, m)

```

## Litter size
```{r}
ls<-read.csv("litter size.csv", sep = ";")
head(ls)

# model diagnostics
ggplot(data = ls, aes(sample = litter.size)) +
  stat_qq() +
  stat_qq_line()

ggplot(ls, aes(x = year, y = litter.size)) +
  geom_point() +
  stat_smooth(method = "lm")
m <- lm(litter.size~year, data = ls)
summary(m)
d <- data.frame(residuals = residuals(m),         # Residuals
                     std_residuals = rstudent(m), # Studentized Residuals
                     fitted = fitted(m),          # Fitted values
                     cooks = cooks.distance(m))   # Cook's D
d <- mutate(d, observation = 1:nrow(d))           # Observation number

# histogram of residuals
ggplot(data = d, aes(x = std_residuals)) +
  geom_histogram(bins = 10)
# Check for Normality: QQ Normality plot of residuals
ggplot(data = d, aes(sample = std_residuals)) +
  stat_qq() +
  stat_qq_line()
# Check heteroscedasticity or constant variance
ggplot(d, aes(x = fitted, y = std_residuals)) +
  geom_point() +
  geom_hline(yintercept = 0)
# Cook's D
ggplot(d, aes(x = observation, y = cooks)) +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_hline(yintercept = 4/nrow(d), linetype = "dashed")
# outliers
d_outliers <- filter(d, cooks > 4/nrow(d))
d_outliers
ls[c(16), ] # correct datapoint
# problem with heteroscedasticity --> nonparametric test

kruskal.test(litter.size~year, data = ls)

lssum<-group_by(ls, year) %>%
  summarise_all(funs(mean, sd, length)) %>%
  mutate(se=sd/sqrt(length))
sd(ls$litter.size)/sqrt(length(ls$litter.size))
lssum

# clean environment
rm(d_outliers, d, m, ls)
```

## litter sex ratio
Sex ratio data are best expressed as proportions (sex ratio = males/(males + females))
deviations are expected to conform to a binomial distribution
from
Wilson, K., & Hardy, I. (2002). Statistical analysis of sex ratios: An introduction. In I. Hardy (Ed.), Sex Ratios: Concepts and Research Methods (pp. 48-92). Cambridge: Cambridge University Press. doi:10.1017/CBO9780511542053.004
```{r}
jsr<-read.csv("juvenile sex ratio.csv", sep = ";")
# quick check
t.test(jsr$male.juveniles, jsr$female.juveniles) 
t.test(jsr$male.juveniles, jsr$female.juveniles, paired = TRUE)

# differ across years?
jsr<-select(jsr, year, sex, juveniles)
head(jsr)

jsrsum<-group_by(jsr, year, sex) %>%
  summarise(sum=sum(juveniles))

# per year
binom.test(c(15,15),30, p=0.5) # 2014
binom.test(c(18,20),38, p=0.5) # 2015
binom.test(c(33,30),63, p=0.5) # 2016

males<-c(15, 18, 33)
females<-c(15, 20, 30)

rows=3
m<-matrix(c(males, females),nrow=rows)
rownames(m) = c("2014", "2015", "2016")      # Naming the rows and
colnames(m) = c("males", "females")          # columns is optional
m

chisq.test(m)

# overall 
binom.test(c(66,65),131, p=0.5)

rm(males, females, m, jsrsum, jsr)
```

## juvenile mortality
does juvenile mortality differ per year?
```{r}
# Juvenile mortality (%)	19.4	50.3	49.4
# emerged juveniles 112, 161, 166

0.194*112 # 2014
0.503*161 # 2015
0.494*166 # 2016

surv<-c((112-22), (161-81), (166-82))
nonsurv<-c(22, 81, 82)

m<-matrix(c(surv, nonsurv),nrow=rows)
rownames(m) = c("2014", "2015", "2016")          # Naming the rows and
colnames(m) = c("survived", "not-survived")   #  columns is optional
m

chisq.test(m)
# df= (row-1)(column-1)
(3-1)*(2-1)

rm(surv, nonsurv, m)
```

## adult lifespan and survival rate
```{r}
# life span
age <- c(1,2,3,4)
certain <-	c(30,	17,	3,	0)
uncertain	<- c(127,	54,	35,	14)
df<-cbind.data.frame(age, certain, uncertain)

ggplot(df, aes(x=age, y = certain)) +
  geom_line() +
  labs(y ="survivors") +
  theme_classic() +
  geom_line(aes(x=age, y=uncertain), linetype=2)

rm(age, certain, uncertain, df)

# survival
#https://www.datacamp.com/community/tutorials/survival-analysis-R
# Load required packages
library(survival)
library(survminer)
library(dplyr)

survival<-read.csv("Aget_survival.csv", sep = ";")
s<-filter(survival, survival!=0)

s<-janitor::clean_names(s)
# age = certain and uncertain with +
# survival = uncertain age not censored
# age_censored (1=certain age, 0=uncertain age)

hist(s$survival)
# Fit survival data using the Kaplan-Meier method
surv_object <- Surv(time = s$survival, event = s$age_censored)
surv_object 

fit <- survfit(surv_object~1, data = s)
summary(fit)
ggsurvplot(fit, data = s, pval = TRUE)

fit1 <- survfit(surv_object~sex, data = s)
summary(fit1)
ggsurvplot(fit1, data = s, pval = TRUE)
```


# AIM 2: comparison life history traits to other mammals 

Where do Barbaries fit with life history traits in comparison to other mammals?
```{r}
# dataset Capellini et al. 2015 ----
data_cap<-read.csv("cappelini.csv", sep = ";") # successfully spread mammals
data_cap_Ag <- data_cap %>% 
  filter(Binomial!="Atlantoxerus_getulus")
head(data_cap)

df.cap<-data_cap_Ag
df.cap$species <- factor(df.cap$family, levels= c("all","sciuridae"),
  labels = c("mammals", "only squirrels"))

df.median<- df.cap %>% group_by(family)
df.median<-  summarise_if(df.median, is.numeric, median, na.rm = TRUE) 
barbaries<-c("A. getulus", 1, 5, 221.1, 43.5, 66, 8.9, 2.9, 1, 344.8, 4.1)
df.median.all<-rbind(df.median, barbaries)
median_long <- gather(df.median.all, trait, value, NoLocs:RLy)
median_long$value<-as.numeric(median_long$value)
median_long <- median_long %>% 
  mutate_if(is.numeric, round, digits = 1)
df_median<-pivot_wider(median_long, names_from = trait, values_from = value ) %>% group_by(family)

df_median$family<-as.factor(df_median$family)

# set color and other parameters
col.teal <- "#1f78b4" #use for crowds
col.ltgn <- "#b2df8a" #use for sparse displace
col.dkgn <- "#33a02c" #use for regular displace
col.lttl <- "#a6cee3"
col.dkrd <- "#ca0020"
# "#D1E5F0" "#FDDBC7"

#set color by group
squirrels <- "#FF7F00"
mammals <- "#377EB8"
obs.vals <- "#4DAF4A" # "#E41A1C"
species.col<- c("#377EB8", "#FF7F00", "#4DAF4A" )

#Create a custom color scale for species
library(RColorBrewer)
myColors <- c("#377EB8","#FF7F00" ) 
names(myColors) <- levels(df.cap$family)
colScale <- scale_colour_manual(name = "family",values = myColors)

graphics.off()
dev.off()

```


ggplot
```{r}


# plot Nolocs
plot.nolocs <- ggplot(df.cap, aes(x=NoLocs, col = family)) + 
  geom_density(alpha=0.4) +
  labs(y ="density" , x ="number of introductions") +
  colScale +
  theme_classic() +
   geom_vline(data = df.median, aes(xintercept=NoLocs, color = family), 
            linetype="dashed") +  # linetype="dashed", size=1 +
  geom_vline(aes(xintercept=median(data_cap$LG[data_cap$Binomial=="Atlantoxerus_getulus"])), # add squirrel line
             color=obs.vals, linetype="dashed", size=1)+
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001))

plot.nolocs

# longevity
plot.lg <- ggplot(df.cap, aes(x=LG, col = family)) + 
  geom_density() +
  labs(y ="" , x ="longevity (years)") +
  colScale +
  theme_classic() +
   geom_vline(data = df.median, aes(xintercept=LG, color = family), 
            linetype="dashed", size=1) +
  geom_vline(aes(xintercept=median(data_cap$LG[data_cap$Binomial=="Atlantoxerus_getulus"])), # add squirrel line
            color=obs.vals, linetype="dashed", size=1)+
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001))
plot.lg

# body mass
plot.bm <- ggplot(df.cap, aes(x=BM, col = family)) + 
  geom_density() +
  labs(y ="" , x ="body mass (g)") +
  colScale +
  theme_classic() +
  xlim(c(0,10000)) +
  geom_vline(aes(xintercept=median(data_cap$BM[data_cap$Binomial=="Atlantoxerus_getulus"])), # add squirrel line
            color=obs.vals, linetype="dashed", size=1) +
  geom_vline(data = df.median, aes(xintercept=BM, color = family), 
            linetype="dashed") +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001))
plot.bm

# Neonatal body mass
plot.nbm <- ggplot(df.cap, aes(x=NBM, col = family)) + 
  geom_density() +
  labs(y ="" , x ="neonatal body mass (g)") +
  colScale +
  theme_classic() +
  xlim(c(0,150)) +
  geom_vline(aes(xintercept=median(data_cap$NBM[data_cap$Binomial=="Atlantoxerus_getulus"])), # add squirrel line
            color=obs.vals, linetype="dashed", size=1) +
  geom_vline(data = df.median, aes(xintercept=NBM, color = family), 
            linetype="dashed") +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001))
plot.nbm

# litter size
plot.ls <- ggplot(df.cap, aes(x=LS, col = family)) + 
  geom_density() +
  labs(y ="density" , x ="litter size") +
  colScale +
  theme_classic() +
  geom_vline(aes(xintercept=median(data_cap$LS[data_cap$Binomial=="Atlantoxerus_getulus"])), # add squirrel line
            color=obs.vals, linetype="dashed", size=1) +
  geom_vline(data = df.median, aes(xintercept=LS, color = family), 
            linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001))
plot.ls


# litters/year
plot.ly <- ggplot(df.cap, aes(x=LY, col = family)) + 
  geom_density() +
  labs(y ="" , x ="litters/year") +
  colScale +
  theme_classic() +
   geom_vline(data = df.median, aes(xintercept=LY, color = family), 
            linetype="dashed", size=1) +
  geom_vline(aes(xintercept=data_cap$LY[data_cap$Binomial=="Atlantoxerus_getulus"]), # add squirrel line
            color=obs.vals, linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001))
 
plot.ly


# age at first birth
plot.afb <- ggplot(df.cap, aes(x=AFB, col = family)) + 
  geom_density() +
  labs(y ="density" , x ="age at first birth (days)") +
  colScale +
  theme_classic() +
   geom_vline(data = df.median, aes(xintercept=AFB, color = family), 
            linetype="dashed", size=1) +
  geom_vline(aes(xintercept=median(data_cap$AFB[data_cap$Binomial=="Atlantoxerus_getulus"])), # add squirrel line
            color=obs.vals, linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001))
plot.afb

# gestation time
plot.gt <- ggplot(df.cap, aes(x=GT, col = family)) + 
  geom_density() +
  labs(y ="density" , x ="gestation time (days)") +
  colScale +
  theme_classic() +
   geom_vline(data = df.median, aes(xintercept=GT, color = family), 
            linetype="dashed", size=1) +
  geom_vline(aes(xintercept=median(data_cap$GT[data_cap$Binomial=="Atlantoxerus_getulus"])), # add squirrel line
            color=obs.vals, linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001))
plot.gt


# weaning age
plot.wa <- ggplot(df.cap, aes(x=WA, col = family)) + 
  geom_density() +
  labs(y ="" , x ="weaning age (days)") +
  colScale +
  theme_classic() +
  geom_vline(data = df.median, aes(xintercept=WA, color = family), 
            linetype="dashed", size=1) +
  geom_vline(aes(xintercept=median(data_cap$WA[data_cap$Binomial=="Atlantoxerus_getulus"])), # add squirrel line
            color=obs.vals, linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001))
plot.wa


# reproductive lifespan
plot.rl <- ggplot(df.cap, aes(x=RLy, col = family)) + 
  geom_density() +
  labs(y ="" , x ="reproductive lifespan (years)") +
  colScale +
  theme_classic() +
   geom_vline(data = df.median, aes(xintercept=RLy, color = family), 
            linetype="dashed", size=1) +
  geom_vline(aes(xintercept=median(data_cap$RLy[data_cap$Binomial=="Atlantoxerus_getulus"])), # add squirrel line
            color=obs.vals, linetype="dashed", size=1) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.001))
plot.rl


plot.compare<-ggarrange(plot.nolocs,plot.lg,  plot.gt, plot.wa, plot.ls, plot.ly, plot.afb, plot.rl,
                   common.legend = TRUE  , legend = "bottom",
                   nrow = 4, ncol = 2, labels = "auto") #plot.nbm, plot.bm
plot.compare
ggexport(plot.compare, filename = "plot.compare.traits.pdf")



```



# AIM3: mcmcGLMM
we examined whether number of founders and/or a fast life history influences invasion success of squirrels using the dataset provided by Bertolino (2009).

Load libraries
```{r setup, include=FALSE}
library(TreeTools)
library(tidyverse)
library(ape)
library(MCMCglmm)
library(brms)
library(phytools)
sessionInfo()
```

## Import data
```{r}
sq_df<-read.csv("bertolino.csv", sep =";") %>%
  filter(Founders!="NA") %>%
  dplyr::select(-ref, -NBM)
sq<-sq_df
head(sq)

sq$Intro<-as.factor(sq$Intro)
sq$Est<-as.factor(sq$Est)
sq$Spread<-as.factor(sq$Spread)
sq$spread<-factor(sq$Spread, levels = c("0", "1"))


# number of times each species is included
#group_by(sq,  Binomial, Spread) %>%
#  dplyr::summarise(n=n())
#group_by(sq,  Binomial) %>%
#  dplyr::summarise(n=n())

# species names (to include in tree)
species<-unique(sq$Binomial)
species
```

## import phylogenetic tree

We used the mammalian supertree from Rolland et al. 2014, see also Lukas & Clutton-brock 2020. 

```{r}
Rollandtree <- phytools::read.newick("Rolland et al tree.txt")
#plot(Rollandtree)
str(Rollandtree)
tree<-Rollandtree
```


## Collinearity of variables
Prepare data
```{r}
## Collinearity of the explanatory variables ----
library(psych) #Calls: pairs.panels
library(car) #Calls: vif
library(plyr) #Calls: rbind.fill
library(usdm)

# global
Z <- cbind(sq$Founders, sq$NoLocs, sq$LG, sq$BM, sq$GT, sq$WA ,
           sq$LS , sq$LY , sq$AFB , sq$RL)
colnames(Z) <- c("f", "nl", "lg","bm","gt","wa","ls","ly", "afb","rl")
pairs.panels(Z) # Generate scatterplots with Pearson correlations
dfz = data.frame(Z) # Data Frame with predictor variables
vif(dfz) # VIF > 3

Z <- cbind(sq$Founders, sq$NoLocs, sq$LG, sq$BM, sq$GT, sq$WA ,
           sq$LS , sq$LY , sq$AFB)  
colnames(Z) <- c("f", "nl", "lg","bm","gt","wa","ls","ly", "afb")
pairs.panels(Z) # Generate scatterplots with Pearson correlations
dfz = data.frame(Z) # Data Frame with predictor variables
vif(dfz) # VIF > 3: remove rl

Z <- cbind(sq$Founders, sq$NoLocs, sq$BM, sq$GT, sq$WA ,
           sq$LS , sq$LY , sq$AFB)
colnames(Z) <- c("f", "nl","bm","gt","wa","ls","ly", "afb")
pairs.panels(Z) # Generate scatterplots with Pearson correlations
dfz = data.frame(Z) # Data Frame with predictor variables
vif(dfz) # VIF > 3: remove lg

Z <- cbind(sq$Founders, sq$NoLocs, sq$GT, sq$WA ,
           sq$LS , sq$LY , sq$AFB)
colnames(Z) <- c("f", "nl","gt","wa","ls","ly", "afb")
pairs.panels(Z) # Generate scatterplots with Pearson correlations
dfz = data.frame(Z) # Data Frame with predictor variables
vif(dfz) # VIF > 3: remove bm

Z <- cbind(sq$Founders, sq$NoLocs, sq$WA ,
           sq$LS , sq$LY , sq$AFB)
colnames(Z) <- c("f", "nl","wa","ls","ly", "afb")
pairs.panels(Z) # Generate scatterplots with Pearson correlations
dfz = data.frame(Z) # Data Frame with predictor variables
vif(dfz) # VIF > 3: remove gt

Z <- cbind(sq$Founders, sq$WA ,
           sq$LS , sq$LY , sq$AFB)
colnames(Z) <- c("f", "wa","ls","ly", "afb")
pairs.panels(Z) # Generate scatterplots with Pearson correlations
dfz = data.frame(Z) # Data Frame with predictor variables
vif(dfz) # VIF > 3: remove nolocs

# check with dependent and random factor
Z.all <- cbind(sq$Founders, sq$WA ,
           sq$LS , sq$LY , sq$AFB, sq$spread, sq$Binomial)
colnames(Z.all) <- c("f", "wa","ls","ly", "afb", "spread", "species")
pairs.panels(Z.all) # Generate scatterplots with Pearson correlations
dfz = data.frame(Z.all) 
vif(dfz) # VIF > 3: removed ls

Z.all <- cbind(sq$Founders, sq$WA ,
               sq$LY , sq$AFB, sq$spread, sq$Binomial)
colnames(Z.all) <- c("f", "wa","ly", "afb", "spread", "species")
pairs.panels(Z.all) # Generate scatterplots with Pearson correlations
dfz = data.frame(Z.all) 
vif(dfz) # VIF > 3

```



Plot
```{r}
# plots
sq$invasive<-factor(sq$spread, labels = c("no", "yes"))
figA<- ggplot(sq, aes(x=factor(invasive), y=Founders)) + 
  geom_boxplot() +
  theme_classic() +
  labs(x= "", y = "number of founders") #+
  geom_point(aes(group = Binomial)) +
  geom_line(aes(group = Binomial),
            alpha = 0.5, colour = "darkgrey")
  
figB<- ggplot(sq, aes(x=factor(invasive), y=WA)) + 
  geom_boxplot() +
  theme_classic() +
  labs(x= "", y = "weaning age (days)") 

figC<- ggplot(sq, aes(x=factor(invasive), y=LY)) + 
  geom_boxplot() +
  theme_classic() +
  labs(x= "invasive", y = "litters/year") 

figD<- ggplot(sq, aes(x=factor(invasive), y=AFB)) + 
  geom_boxplot() +
  theme_classic() +
  labs(x= "invasive", y = "age at first birth (days)")

library(ggpubr) 
plot.squirrels<-ggarrange(figA, figB, figC, figD, labels = "auto")
ggexport(plot.squirrels, filename = "plot.squirrels.pdf")
```


## Phylogenetic mixed model

standardize values
```{r}

st<-sq %>% mutate_if(is.numeric, scale)
head(st)

```


Phylogenetic tree doesn't include "Callosciurus_erythreus" & "Sciurus_aberti",
therefore exclude resulting in 11 species instead of 13 for our analysis. 

```{r}
st$Binomial<-as.factor(st$Binomial)
unique(st$Binomial)
st<- st %>% filter(!Binomial %in% c("Callosciurus_erythreus", "Sciurus_aberti")) %>% droplevels()
unique(st$Binomial)


```


Only include the species that are present in Bertolino's dataset.
```{r}
pruned.tree<-drop.tip(Rollandtree, setdiff(Rollandtree$tip.label, species)) 
str(pruned.tree)
plot.phylo(pruned.tree)
pruned.tree$tip.label
```


Reorder data according to tips
```{r}
species<-pruned.tree$tip.label
st_tree_select<-dplyr::select(st, Phylo, Binomial,spread, Founders, WA, LY, AFB) 
st_tree_select$Binomial

traits<-  st_tree_select %>% arrange(factor(Binomial, levels = species))
traits$Binomial
```

## repeated measures brms

```{r}
A <- ape::vcv.phylo(pruned.tree) # covariance matrix
```

```{r}
# get mean of repeated measures
st_select_mean<-traits
st_select_mean$Binomial
st_select_mean$mean_founders <- 
  with(traits, sapply(split(Founders, Binomial), mean)[Binomial])
st_select_mean$mean_WA <- 
  with(traits, sapply(split(WA, Binomial), mean)[Binomial])
#st_select_mean$mean_LS <- 
#  with(traits, sapply(split(LS, Binomial), mean)[Binomial])
st_select_mean$mean_LY <- 
  with(traits, sapply(split(LY, Binomial), mean)[Binomial])
st_select_mean$mean_AFB <- 
  with(traits, sapply(split(AFB, Binomial), mean)[Binomial])
st_select_mean$spread<-as.numeric(st_select_mean$spread)
head(st_select_mean)
```

get prior
```{r}
get_prior(spread ~ mean_founders + mean_WA + mean_LY + mean_AFB + 
            (1|gr(Binomial, cov = A))+ + (1|Phylo), 
          data = st_select_mean, 
          family = binomial())
get_prior(spread ~ mean_founders + mean_WA + mean_LY + mean_AFB + 
            (1|gr(Binomial, cov = A))+ + (1|Phylo), 
          data = st_select_mean, 
          family = bernoulli())
```

model
```{r}
## model
model_repeat <- brm(
  spread ~ mean_founders + mean_WA + mean_LY + mean_AFB + 
    (1|gr(Binomial, cov = A))+ (1|Phylo), 
  data = st_select_mean, 
  family = bernoulli(), 
  data2 = list(A = A),
  prior = c(
    prior(student_t(3, 0, 2.5))), 
  sample_prior = TRUE, chains = 3, cores = 2, # results
  iter = 10000, warmup = 1500,
  control = list(adapt_delta = 0.99)
)

# checkout results: parameter settings
#sample_prior = TRUE, chains = 2, cores = 2, # checkout results
#  iter = 4000, warmup = 1000,
#  control = list(adapt_delta = 0.99)


pairs(model_repeat)
summary(model_repeat, waic = TRUE)
head(predict(model_repeat)) # predict responses based on the fitted model
plot(conditional_effects(model_repeat), ask = FALSE) # plot conditional effects for each predictor

# investigate model fit
loo(model_repeat) 
pp_check(model_repeat) 


# get variables
#get_variables(model_repeat)
```



## add variability of the independent variables (predictor) within species
```{r}
# as only founders change within species, we only have to add founders
st_select_mean$within_founders <- st_select_mean$Founders - st_select_mean$mean_founders

model_repeat_within <- brm(
  spread ~ within_founders + mean_founders + mean_WA +  mean_LY + mean_AFB + 
    (1|gr(Binomial, cov = A))+ (1|Phylo), 
  data = st_select_mean, 
  family = bernoulli(), 
  data2 = list(A = A),
  #prior = c(
  #  prior(student_t(3, 0, 2.5))),
  sample_prior = TRUE, chains = 3, cores = 2, #  results
  iter = 10000, warmup = 1500,
  control = list(adapt_delta = 0.999, max_treedepth = 15),
  save_all_pars = TRUE
) 

 sample_prior = TRUE, chains = 2, cores = 2, # checkout results
  iter = 4000, warmup = 1000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  save_all_pars = TRUE

# optional
  sample_prior = TRUE, chains = 3, cores = 2, 
  iter = 1500000, warmup = 500000, thin = 10,
  control = list(adapt_delta = 0.999, max_treedepth = 15)
)

summary(model_repeat_within)
plot(model_repeat_within)
pairs(model_repeat_within)
summary(model_repeat_within, waic = TRUE)
head(predict(model_repeat_within)) # predict responses based on the fitted model
plot(conditional_effects(model_repeat_within), ask = FALSE) # plot conditional effects for each predictor

# investigate model fit
loo(model_repeat_within) 
pp_check(model_repeat_within) 
```


## model selection
```{r}
loo(model_repeat_within, model_repeat) # model repeat lowest looic
loo_compare(model_repeat_within, model_repeat, criterion = "waic")
waic(model_repeat_within, model_repeat) # loo better for the models
```

## plot posterior distribution
```{r}
plot.model<- plot(model_repeat)
ggexport(plot.model, filename = "plot.phylogenetic.model.pdf")
# geom_vline(x=0, linetype = "dashed")
     #pars = "^b_", combo = c("dens", "trace")
```


## explanation model output
Solutions for warning messages
    - control = list(max_treedepth = 15) # <x> a positive integer  > default of 10
    - control = list(adapt_delta = 0.99) # <x> > default of 0.8 (require 0<delta<1)
    
Rhat: If Rhat is considerably greater than 1 (i.e., > 1.1), the chains have not yet 
converged and it is necessary to run more iterations and/or set stronger priors.

(1|gr(Binomial, cov = A)) : accounts for phylogenetic relationship between species
(1|Phylo) : accounts for any specific effect that would be independent of the phylogenetic relationship between species (e.g., environmental or niche effects).

There is a significant effect when credibility interval does not contain 0. 

Investigate model fit
 - loo(model_repeat); leave-one-out cross-validation for direct numerical comparison of model fit, smaller values of loo indicate better fit
 - pp_check(model_repeat); posterior predictive check.